Choosing Color Palettes for Statistical Graphics

### 2006-10-13 trunk/vcd/paper/Color/draft.Rnw
added some references for colors
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@589 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\Abstract{
  Statistical graphics are often augmented by the use of color
  coding information contained in some variable. When this involves
  the shading of areas (and not only points or lines)---e.g., as in barplots, pie charts,
  mosaic displays or heatmaps---it is important that the colors are perceptually based
  and do not introduce optical illusions or systematic bias. Here, we discuss
  how the perceptually-based HCL color space can be used for deriving suitable
  color palettes for coding categorical data (qualitative palettes) and numerical
  variables (sequential and diverging palettes).
}

@@

In this section, we show a few examples for the various types of palettes applied
to statistical graphics. The first illustration is a visualization of a bivariate
density estimation for the Old Faithful geyser eruptions data. Figure~\ref{fig:bkde}
shows heatmaps of a bivariate kernel density estimate of waiting times between
and duration of geyser eruptions in Yellowstone National Park. Both use a sequential
palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
with $L \in [30, 90]$ and $p_2 = 2$. The sequential palette in the left panel uses only gray
colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$.

@@

To illustrate qualitative palettes, data from the 2005 election for the German parliament `Bundestag'
are employed. In that election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party `Die Gruenen' and a purple 
hue $H = 300$ for the leftist party `Die Linke'. All colors use the same chroma $C = 60$ and
luminance $L = 75$. The pie chart clearly shows that neither the governing coalition of SPD
and Gruene nor the opposition of CDU and FDP could assemble a majority. Given that no party
would enter a coalition with the leftists, this lead to a big coalition of CDU and SPD.
Figure~\ref{fig:votes} shows the distribution of votes in this election stratified by 
province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 Western provinces, then for the 6 Eastern provinces. Clearly,
the SPD performed better in the North and the CDU better in the South; furthermore,
Die Linke performed particularly well in the Eastern provinces and the Saarland.

>>>>

\Abstract{
  Statistical graphics are often augmented by the use of color
  coding information contained in some variable. When this involves
  the shading of areas (and not only points or lines)---e.g., as in bar plots, pie charts,
  mosaic displays or heatmaps---it is important that the colors are perceptually based
  and do not introduce optical illusions or systematic bias. Here, we discuss
  how the perceptually-based Hue-Chroma-Luminance (HCL) color space can be used for deriving suitable
  color palettes for coding categorical data (qualitative palettes) and numerical
  variables (sequential and diverging palettes).
}

@@

In this section, we show a few examples for the various types of palettes applied
to statistical graphics. The first illustration is a visualization of a bivariate
density estimation for the Old Faithful geyser eruptions data \citep{vcd:Azzalini+Bowman:1990}.
Figure~\ref{fig:bkde}
shows heatmaps of a bivariate kernel density estimate of waiting times between
and duration of geyser eruptions in Yellowstone National Park. Both use a sequential
palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
with $L \in [30, 90]$ and $p_2 = 2$. The sequential palette in the left panel uses only gray
colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$.

@@

To illustrate qualitative palettes, data from the 2005 election for the German parliament `Bundestag'
are employed. In that election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party `Die Gruenen' and a purple 
hue $H = 300$ for the leftist party `Die Linke'. All colors use the same chroma $C = 60$ and
luminance $L = 75$. The pie chart clearly shows that neither the governing coalition of SPD
and Gruene nor the opposition of CDU and FDP could assemble a majority. Given that no party
would enter a coalition with the leftists, this lead to a big coalition of CDU and SPD.
Figure~\ref{fig:votes} shows the distribution of votes in this election stratified by 
province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 Western provinces, then for the 6 Eastern provinces. Clearly,
the SPD performed better in the North and the CDU better in the South; furthermore,
Die Linke performed particularly well in the Eastern provinces and in Saarland.
